---
title: "Bellabet Case Study: Smart Device Analysis"
author: Anh Phuong Dieu Nguyen
date: May 2026
output: html_notebook
---

```{r}
## Load libraries
library(tidyverse)
library(lubridate)
library(janitor)
library(ggplot2)
```

```{r}
## Load data
sleep <- read_csv("sleepDay_merged.csv",show_col_types = FALSE)
weight <- read_csv("weightLogInfo_merged.csv",show_col_types = FALSE)
activity_daily <- read_csv("dailyActivity_merged.csv",show_col_types = FALSE)
calories_hourly <- read_csv("hourlyCalories_merged.csv",show_col_types = FALSE)
steps_hourly <- read_csv("hourlySteps_merged.csv",show_col_types = FALSE)
intensity_hourly <- read_csv("hourlyIntensities_merged.csv",show_col_types = FALSE)

```

```{r}

## Standardize column names
sleep <- clean_names(sleep)
weight <- clean_names(weight)
activity_daily <- clean_names(activity_daily)
calories_hourly <- clean_names(calories_hourly)
steps_hourly <- clean_names(steps_hourly)
intensity_hourly <- clean_names(intensity_hourly)

## Check data type
all <- list(sleep = sleep, weight = weight, activity_daily = activity_daily, calories_hourly = calories_hourly, steps_hourly = steps_hourly, intensity_hourly = intensity_hourly)

lapply(all, function(x) sapply(x, class))

## Convert date columns
sleep <- sleep |> mutate(sleep_day = as.Date(mdy_hms(sleep_day)))
weight <- weight |> mutate(date = as.Date(mdy_hms(date)))
activity_daily <- activity_daily |> mutate(activity_date = as.Date(mdy(activity_date)))
steps_hourly <- steps_hourly |> mutate(activity_hour =  mdy_hms(activity_hour))
intensity_hourly <- intensity_hourly |> mutate(activity_hour =  mdy_hms(activity_hour))
calories_hourly <- calories_hourly |> mutate(activity_hour =  mdy_hms(activity_hour))

## Check null
lapply(all, function(x) sum(is.na(x)))
colSums(is.na(weight))

## Remove column having many missing values
weight <- weight |> select(-fat)

## Remove duplicated values
sleep <- distinct(sleep)
weight <- distinct(weight)
activity_daily <- distinct(activity_daily)
calories_hourly <- distinct(calories_hourly)
intensity_lhoury <- distinct(intensity_hourly)
steps_hourly <- distinct(steps_hourly)

## Remove unnecessary columns
weight <- weight |> select(-is_manual_report, -log_id)

```

```{r}
## Feature engineering
activity_daily <- activity_daily |>
  mutate(wday = wday(activity_date, label = TRUE),
         activity_level = case_when(
           total_steps < 5000 ~ "Sedentary",
           total_steps < 7500 & total_steps > 5000 ~ "Low Active",
           total_steps < 10000 & total_steps > 7500 ~ "Moderately Active",
           total_steps > 10000 ~ "Active"
         ),
        active_minutes = very_active_minutes + fairly_active_minutes + lightly_active_minutes,
         )

steps_hourly <- steps_hourly |>
  mutate(hour = hour(activity_hour))
calories_hourly <- calories_hourly |>
  mutate(hour = hour(activity_hour))
intensity_hourly <- intensity_hourly |>
  mutate(hour = hour(activity_hour))
  
```

```{r}
## Join necesssary tables
daily <- activity_daily |> 
  left_join(sleep, by = c("id","activity_date"="sleep_day" )) |>
  mutate(id = as.factor(id))
hourly <- calories_hourly |> 
  left_join(steps_hourly, by=c("id","activity_hour","hour")) |>
  left_join(intensity_hourly, by=c("id","activity_hour","hour")) |>
  mutate(id = as.factor(id))
```

```{r}
## Exploratory Data Analysis
#1. Average sleep per  user

summary <- daily |>
  group_by(id) |>
  summarise(
    avg_sleep = mean(total_minutes_asleep, na.rm = TRUE),
    avg_calories = mean(calories, na.rm=TRUE),
    avg_steps = mean(total_steps, na.rm = TRUE),
    .groups = "drop"
  )
avg_total_sleep <- mean(summary$avg_sleep, na.rm=TRUE)
avg_total_sleep

sleep_plot <- summary |> 
  filter(!is.na(avg_sleep))|> 
  arrange(avg_sleep) |> 
  ggplot(aes(x=reorder(id,avg_sleep),y=avg_sleep)) + 
  geom_col(fill="#7CA37C") + 
  theme(axis.text.y = element_blank()) +
  coord_flip() +
  geom_hline(yintercept=avg_total_sleep, linewidth = 1, color =  "#bb0a1e") +
  annotate("text", x=3, y=avg_total_sleep, label="Average (377.44 minutes ~ 6.3 hours/day)", size = 3, hjust = -0.1, vjust = -0.5, color="#bb0a1e") +
  geom_hline(yintercept = 420, linewidth = 1) +
  annotate("text", x = 2, y = 420, label = "Sleep Goal (420 minutes ~ 7 hours/day)", size = 3, hjust = -0.1, vjust = -0.5) +
  labs(
    title = "Average sleep duration per day by User",
    x = "Users",
    y = "Average minutes asleep"
  ) 
```

```{r}
#2. Average steps per user
avg_total_steps <- mean(summary$avg_steps)
steps_plot <- summary |> 
  arrange(avg_steps) |>
  ggplot(aes(x=reorder(id, avg_steps),y=avg_steps)) +
  geom_col(fill="#4169E1") +
  theme(axis.text.y=element_blank()) +
  coord_flip() +
  geom_hline(yintercept=avg_total_steps, linewidth = 1, color =  "#bb0a1e") +
  annotate("text", x=3.5, y=round(avg_total_steps,0), label="Average steps (7520 steps/day)", size = 3, hjust = -0.1, vjust = -0.5, color="#bb0a1e") +
  geom_hline(yintercept = 10000, linewidth =1) +
  annotate("text",x=2,y=10000, label = "Steps Goal (10000 steps/day)", size = 3, hjust = -0.1, vjust = -0.5) +
  labs(
    title = "Avarage steps per day by User",
    x = "Users",
    y = "Average number of steps"
  )
```

```{r}
#3. Relationship between number of steps and calories burned
steps_calories <- daily |> 
  ggplot(
    aes(x=total_steps,y=calories)
  ) +
  geom_jitter(colour = "#4169E1",alpha = 0.6 ) + 
  stat_smooth(
    method ="lm",
    formula = y ~ x,
    geom = "smooth",
    color = "black"
  ) +
  labs(
    title = "Relationship between number of steps and calories burned",
    x = "Number of steps per day",
    y = "Calories burned per day "
  )
  
```

```{r}
#4. Proportion of activity level
prop_active <- daily |>
  group_by(activity_level) |> 
  summarise(count = n(),
  .groups = "drop") |> 
  mutate(prop = round(count/sum(count)*100,2))|>
  mutate(highlight = prop == max(prop)) |> 
  ggplot(aes(activity_level,prop, fill= highlight)) +
  geom_col(show.legend = FALSE) +
  scale_fill_manual(
    values = c("TRUE" =  "#FDB44E", "FALSE" = "#f8e472",
    guide = "none")
  ) + 
  geom_text(aes(label = paste0(prop, "%")), vjust = -0.5, size = 2.5) +
  labs(
    title = "Proportion of different activily levels across all users",
    x = "Activity level",
    y = "Proportion"
  )
```

```{r}
#5. Relationship between number of steps and sleep duration
steps_sleep <- daily |>
  filter(!is.na(c(total_minutes_asleep))) |> 
  ggplot(aes(x=total_steps, y=total_minutes_asleep)) +
  geom_point(colour = "#7ca37c",alpha = 0.6 ) +
  stat_smooth(
    method ="lm",
    formula = y ~ x,
    geom = "smooth",
    color = "black"
  ) +
  labs(
    title = "Relationship between number of steps and sleep duration",
    x="Steps per day",
    y="Sleep duration"
  )
```

```{r}
#6. Relationship between sedentary level and sleep duration and sleep quality
sedentary_sleep <- daily |>
  filter(!is.na(c(total_minutes_asleep))) |> 
  ggplot(aes(x=sedentary_minutes, y=total_minutes_asleep)) +
  geom_point(colour = "#FDB44E",alpha = 0.6 ) +
  stat_smooth(
    method ="lm",
    formula = y ~ x,
    geom = "smooth",
    color = "black"
  ) +
  labs(
    title = "Relationship between sedentary level and sleep duration",
    x="Sedentary minutes",
    y="Sleep duration"
  )
```

```{r}
#7. BMI information
avg_bmi <- weight |>
  group_by(id) |>
  summarise(
    avg_bmi = mean(bmi)
  ) |>
  mutate(
    healthy = case_when(
      avg_bmi < 18.5 ~ "Underweight",
      avg_bmi <= 24.9 & avg_bmi >= 18.5 ~ "Healthy",
      avg_bmi > 24.9& avg_bmi <= 29.9 ~ "Overweight",
      avg_bmi > 29.9 ~ "Obesity"
      )
    ) |> 
  group_by(healthy) |>
  summarize(healthy_count = n()) |>
  mutate(proportion = round(healthy_count/sum(healthy_count)*100,2)) |> 
  ggplot(aes(x=healthy, y = proportion, fill = healthy)) +  
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(
    values = c("Underweight" = "#bb0a1e",
    "Healthy" = "#7ca37c",
    "Overweight" = "#bb0a1e",
    "Obesity" = "#bb0a1e")) +
  geom_text(aes(label = paste0(proportion, "%")), vjust = -0.5, size = 2.5) +
  labs(
    title = "Proportion of BMI categories across all users",
    x = "BMI index",
    y = "Proportion of Users"
  )
  
```

```{r}
## Recommendations 
#1. Number of steps in a week
steps_summary1 <- daily |>
  group_by(wday) |>
  summarise(
    sum_steps1 = sum(total_steps),
    .groups = "drop"
  ) |>
  mutate(
    highlight = sum_steps1 == max(sum_steps1)
  )

steps_week <- 
  ggplot(steps_summary1, aes(wday,sum_steps1, fill = highlight)) +
  geom_col() +
  scale_fill_manual(
    values = c("TRUE" = "#0A2472", "FALSE" = "#4169E1"),
    guide = "none"
  ) +
  labs(
    title = "Distribution of number of steps during a week",
    y = "Number of steps",
    x = "Day of Week"
  )
```

```{r}
#2. Number of steps in a day
steps_summary2 <- hourly |>
  group_by(hour) |>
  summarise(
    sum_steps2 = sum(step_total),
    .groups = "drop"
  ) |>
  mutate(
    highlight = sum_steps2  == max(sum_steps2)
  )

steps_day <- 
  ggplot(steps_summary2, aes(hour,sum_steps2, fill = highlight)) +
  geom_col() +
  scale_fill_manual(
    values = c("TRUE" = "#06402B", "FALSE" = "#7CA37C"),
    guide = "none"
  ) +
  labs(
    title = "Distribution of number of steps during a day",
    y = "Number of steps",
    x = "Hour"
  ) +
  scale_x_discrete(
    limits = as.character(0:24)
  )
```

# 1. Introduction

Bellabeat is a succesfull small high-tech manufacturer of health-focused products for women, with beautifully designed technology along with data collection on activity, sleep, stress and reproductive health to empower women with knowledge on their own health and habits. Therefore, Bellabeat is believed to have the potential to become a large player in the global smart device market.

# 2. Business Task

This analysis project will focus on smart fitness data to gain insight how customers are using their smart devices, and in a result, help guide the marketing strategy for the company and provide high-level recommendations for Bellabeat's marketing strategy.

Key Questions:

-   What are some trends in smart device usage?

-   How could these trends apply to Bellabeat customers?

-   How could these trends help influence Bellabeat market strategy?

Stakeholders:

-   **Urška Sršen**: Co-founder and Chief Creative Officer

-   **Sando Mur**: Mathematician and Key member of Bellabeat executive team

-   **Bellabeat marketing analytics team**

Assessment used:

-   Activity level:

    -   \<5000 steps/day: Inactive/Sedentary

    -   5000-7500: Low active

    -   7500-10000: Moderately active

    -   \>10000: Active

-   BMI categories:

    -   \<18.5: Underweight

    -   18.5 - 24.9: Healthy weight

    -   25 - 29.9: Overweight

    -   \>30.0: Obesity

# 3. Data Source

Fitbit Fitness Tracker Data is collected by a survey via Amazon Mechanical Turk between March 12, 2016 to May 12, 2016, containing personal fitness tracker data from 30 Fitbit users, who were consented to the submission of personal trackers data. This data includes information about daily activity, steps, and heart rates that can be used to explore user's habit

Data tables used:

-   sleepDay: sleep duration in minutes (in bed and actually asleep)

-   hourlyCalories: hourly number of calories burned per hour per day

-   hourlyIntensities: hourly total number of intensity and average intensity per hour per day

-   hourlySteps: hourly number of steps per hour per day

-   dailyActivity: combined information on total number of steps, calories, activity level by minutes per day

-   weightLogInfo: weight (in kg and pound) and BMI index recorded on different dates

Sampling bias: there were only 33 Fitbit users conducted the survey, thus may not fully represent the community of smart devices user.

ROCCC:

-   Reliable: Partially

-   Original: Yes

-   Comprehensive: Partially

-   Current: No

-   Cited: Yes

# 4. Data Cleaning

## a. Tool Description

Tool used: R

Library used:

-   tidyverse: data manipulation, cleaning and transformation

-   lubridate: convert date/time formats and extract components like year, month, day, or hour

-   janitor: data cleaning with clean columns names and remove duplicate values

-   ggplot2: create charts to visualize information

## b. Cleaning and Transformation Steps

Cleaning:

-   Standardized column names using `clean_names()`.
-   Checked column data types and converted date and time columns into proper datetime formats.
-   Identified missing values using `is.na()` and `colSums()`.
-   Removed the `fat` column due to excessive missing values.
-   Removed duplicate records using `distinct()`.
-   Dropped unnecessary columns (`is_manual_report`, `log_id`).

Feature engineering:

-   Created a weekday (`wday`) feature from activity dates for weekly activity analysis.
-   Categorized users into activity levels based on total daily steps.
-   Calculated total active minutes by combining very active, fairly active, and lightly active minutes.
-   Extracted hourly information from datetime columns for hourly trend analysis.

Joining tables:

-   daily: left join daily activity data with sleep data

-   hourly: join all hourly data (calories, steps and intensities)

# 5. Data Analysis

## a. Smart device user's BMI overview

```{r}
print(avg_bmi)
```

Around 2/3 of smart devices users in this survey are overweight and/or obese while none of them are underweight. This reflects the difference in the level of health awareness and smart device need towards the underweight and overweight issues, in which the overweight might be more aware and more likely to use health-track smart devices. his can be a great opportunity for marketing teams to tackle this gap in the market and target a new group of customers.

## b. Average sleep and steps per day

```{r}
print(steps_plot)
```

As can be seen from the plots above, both average of sleep and steps are under the recommended level, with average of around 6.3 hours of sleep and around 7520 steps per day, compared with ideal 7 hours of sleep and 10000 steps. In details, less than 50% of users met the goal of 7 hours of sleep, and around a quarter of users reached 10000 steps a day.

While the users in this dataset demonstrate moderately healthy behaviors, with an average of 6.3 hours of sleep and over 7500 steps per day, these figures remain below recommended health guidelines. Therefore, Bellabeat has the potential to introduce new data-driven wellness initiatives and personalized recommendations aimed at helping users improve both sleep quality and daily physical activity.

## c. Number of steps vs. Calories burned

```{r}
print(steps_calories)
```

A positive regression line represents a strong positive relation between the number of steps and the amount of calories burned, indicating that step count is a reliable indicator of energy expenditure.

## d. Number of steps vs. Sleep duration

```{r}
print(steps_sleep)
```

There is a really weak correlation between the number of steps and sleep duration per day, as the regression line appears to be almost flat. Therefore, the number of steps per day is not a good indicator for the sleep quality.

## e. Sedentary duration vs. Sleep duration

```{r}
print(sedentary_sleep)
```

It appears to have a strong negative relationship between sedentary minutes and sleep duration, in which the users having more sedentary time in the day will tend to have a shorter sleep duration. Hence, it is opportunity for Bellabeat to pay attention to focus on features and recommendations that encourage users to reduce sedentary behavior and improve sleep habits.

## f. Distribution of steps during a week

```{r}
print(steps_week)
```

Tuesday is the day having the highest accumulated number of steps, followed by Wednesday and Thursday, while Sunday and Monday are the days with lowest number.

## g. Distribution of steps during a day

```{r}
print(steps_day)
```

5:00 pm appears to be the peak hour in the day with highest number of steps. The distribution of number of steps during a day is also bimodal, with the time period of early afternoon (11:00 am to 1:00 pm) and later afternoon (4:00 pm to 6:00 pm) having pretty high accumulated number of steps.

## h. Key findings summary

-   Around two-thirds of users are overweight or obese, indicating a missed targeted audience - the underweight
-   Average sleep (6.3 hours) and daily steps (7,520) are below recommended health guidelines of WHO
-   Less than half of users meet sleep recommendations, and only about 25% achieve 10,000 daily steps.
-   Daily steps are strongly positively related to calories burned.
-   Step count has little relationship with sleep duration, while there is sedentary time is associated with shorter sleep duration.
-   Activity levels are highest on Tuesday and lowest on Sunday and Monday.
-   Step activity peaks around lunch (11:00 am to 1:00 pm) and late afternoon (4:00 pm to 6:000 pm), especially at 5:00 pm.

# 6. Recommendation

## a. New target expansion - Underweight customers

Bellabeat's marketing campaigns should be developed to promote wellness features specifically designed for underweight users, focusing on:

-   Healthy weight gain

-   Nutrition tracking

-   Wellness monitoring

-   Strength and recovery improvement

## b. Personalized reminders for activity and steps goal-tracking features

Bellabeat should implement personalized reminders and adaptive step goals that:

-   Encourage movement during peak activity hours (Tuesday, from 4:00 pm to 6:00 pm)

-   Send motivational nudges on low-activity days (Sunday and Monday)

-   Adjust goals based on user history and activity level

## c. Enhance Sleep Tracking Through Sedentary Behavior Analysis

A marketing campaign can be launched for its sleep monitoring system by integrating sedentary-time tracking and sleep quality analysis, with potential features can be:

-   Alerts after prolonged inactivity

-   Evening movement reminders

-   Insights connecting inactivity patterns to poor sleep

# 7. Conclusions

Through the insights drawn from this analysis of Fitbit smart devices users, Bellabeat can develop the potential features to develop the marketing strategy along with targeting the untouched group of customers to differentiate themselves as a competitor in the smart devices market.
